Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-08-11 Manuscript/MPESubmission/Revision/response.tex

% -----------------
% Comment 4
% -----------------
\item{\textcolor{blue}{\textbf{Introduction:} Needs stronger emphasis on what specifically is unique about this study compared to what the many other studies on missing data have done and found.
Clearly outlining the questions in this study that are left open from previous research is important because some of the findings in this study suggest that there may be more error associated with missing data than previous research has suggested.}}

We have increased our emphasis on novelty throughout the introduction by more clearly stating the difference between the present study and the literature and more clearly outlining the question of this study, as follows:

% 0(before this section)-Missing data can be a problem
% 1-And until now, not many studies

% NC: Line numbers? TG: I'll do all line numbering once I've incorporated all the edits to the manuscript.
\textit{Until now, few attempts have been made to study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework (i.e. using both molecular and morphological data; e.g. Wiens et al., 2005; Manos et al., 2007; Pattinson et al., 2014).
% 2-These studies have done X and Y to see if missing data's a problem
These studies have assessed the effect of missing data on topology by either (1) comparing a dataset with missing data to subsets without missing data (Wiens et al. 2005); or (2) removing both molecular and some morphological data from living taxa to create artificial fossils (Manos et al., 2007; Pattinson et al., 2014).
% 3-They show that it's not a problem
Both approaches have shown that missing data are not a major problem and should not be an obstacle to combining both living and fossil species in the same phylogenies.
% 4-But they've done it a certain way
% NC: This is still a problem (it's ok up to 4). WHY is an empirical-only approach limiting? WHY is it limiting to generate missing data using fossil patterns?
These studies, however, were limited by because of their empirical approach limiting the conclusions of the studies to similar scenarios. %TG: or scenari? Reviewer 1 is picky with these things
Additionally, Wiens et al. (2005) didn't include fossil taxa in their analyse thus not allowing to test the placement of fossil taxa; and Manos et al. (2007) and Pattinson et al. (2014) used only patters of the fossil record to generate missing data thus not allowing to test the effect of missing data for other patterns observed for living taxa (e.g. the complete absence of morphological data; Guillerme & Cooper 2015).
% TG: Does that sound better?
Besides, Manos et al. (2007) and Pattinson et al. (2014), mainly focused on the palaeontological aspect of the question.} lines @@@.
% NC: I don't like all these i.e.s here. Why not be specific in the text rather than throwing everything into brackets? If you're only citing manos and patterson then just say what they did. TG: no more i.es then ;).
% (next §)5-We're doing it this way

\textit{In this study, we propose a theoretical assessment of the effect of missing data in the Total Evidence method by removing living taxa with morphological data, fossil data, all data for certain characters and the combination of these three aspects.
This is an advance on previous studies because we use large scale simulations and and analyse the effect of three distinct aspect of missing data thus focusing on both neontological and palaeontological parts of the matrix.
Besides, we test the effect of missing data by measuring two crucial aspects of topology in both Maximum Likelihood and Bayesian phylogenies: (i) the conservation of clades (based on the Robinson-Foulds distance; Robinson and Foulds, 1981) and (ii) the displacement of wild-card taxa (based on the Triplets distance; Critchlow et al., 1996) rather than just the clade conservation or support (cf. Wiens et al., 2005; Pattinson et al., 2014).} lines @@@
% NC: Again how is this an advance. I guess it's related to this sentence from below:
% (cf. Wiens et al., 2005; Pattinson et al., 2014 measuring respectively only the clades support or the number of shared clades).

% TG: I actually rewrote this bit as well:
% NC: line numbers?
% \textit{We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the resulting tree topology.
% We inferred the topology from the matrices using both Maximum Likelihood and Bayesian Inference methods and measured the differences in topology using two different topological distance metrics as proxies for clade conservation (based on the Robinson-Foulds distance; Robinson and Foulds, 1981) and for wild-card taxa placement (based on the Triplets distance; Critchlow et al., 1996).
% Thus, in the present study, we propose a new approach to the missing-data question, focusing on both palaeontological and neontological data (rather than just palaeontological data), using only simulated data and measuring the effects of missing data on two aspects of topology (cf. Wiens et al., 2005; Pattinson et al., 2014 measuring respectively only the clades support or the number of shared clades).} lines @@@
% NC: This paragraph seems to be repeating what you've already said above but in slightly more detail. Might make more sense to combine this paragraph and the one above it? TG: Ok, I've combined them

@@

% -----------------
% Comment 5
% -----------------
\item{\textcolor{blue}{\textbf{Abbreviation:} One suggestion is to consider abbreviating Maximum Likelihood and Bayesian Inference as ML and BI respectively early in the methods and use throughout.}}

We understand the reviewer's concern.
Nonetheless, we would prefer not abbreviate these terms. 
Although they might make sentences where Maximum Likelihood and/or Bayesian Inference are stated more concise, we think spelling out the terms makes our paper clearer and does not ask the reader to look back for the actual meaning of the abbreviation (Berlin 2013; Radiology; \url{http://pubs.rsna.org/doi/abs/10.1148/radiol.12121776}).
Additionally this could lead to a confusion between ML (Maximum Likelihood) and $M_L$ (Proportion of living taxa with no morphological data).
However, we can change this at a later stage if necessary.

% -----------------
% Comment 6
% -----------------
\item{\textcolor{blue}{\textbf{Methodology - True tree:} I suggest it warrants comparing at least the 'best' tree to the 'true' tree.}}

We have now compared both the ``best'' tree and the ``missing-data'' trees to the ``true'' tree as suggested by both reviewers.
The results are available in appendix B; figure B1 and appendix C; figures C6 and C7.
The ability to recover the ``true'' tree is poor for both methods and using both metrics.
When comparing the ``true'' tree to the ``best'' trees, in Maximum Likelihood the 50\% CI lays between approximately 0.2 and 0.3 for the Normalised Robinson-Foulds and Triplets metric; and, in Bayesian consensus tree  the 50\% CI lays between approximately 0.3 and 0.4 for the Normalised Robinson-Foulds metric and between 0 and 0.2 for the Normalised Triplets metric). % NC: What are these results comparing? All missing data trees  to the true tree? True trees to best trees? Clarify. TG: is that ok?
This may be related to the size of our matrices or to the ability of the algorithms to generate the ``complete matrix'' but also to our decision to put our minimum median bootstrap acceptance at 50 rather than higher values (line @@@ in the former manuscript and @@@ in our revision).
When comparing the ``true'' tree to the ``missing data'' trees, the results are \textit{scaled down} due to the low scores of the ``true'' \textit{vs.} ``best'' trees scores.
However, we can observe the same patterns between the methods as when using the ``best'' tree (i.e. Bayesian always outperforms ML in clade conservation but does poorly in wild-card taxa placement).
An peculiar feature is that the Normalised Robinson-Foulds metric actually \textit{increases} when we compare the ``missing data'' trees to the ``true'' tree. % NC: specify what is being compared, e.g. "when we compare the missing data trees to the ``true'' trees."" 
This is because the difference between the ``best'' and the ``true'' tree is high, making trees with more data more different than those with lots of missing data.
When data are missing, however, more clades are unresolved and thus closer to clades from the ``true'' trees.

@@

\textit{The Bayesian consensus trees, however, perform badly for the Normalised Triplets metric: some parameter combinations, especially when the $M_F$ parameter reaches 75\% missing data, lead to negative values (figure 5).
A Normalised Triplets metric value below 0 means that the placement of some taxa is worse than expected by just randomly placing this taxon in the tree.
This can be interpreted as the absence of comparable triplets between some of the ``missing data" trees and ``best'' trees.
Even if clades are conserved (figure 5), the resolution within them can be poor to non-existent when a lot of data are missing (i.e. 75\%).
In such cases, the fossil taxa are likely to be placed in any of the clades that share the most characters.
These results are in agreement with previous studies that have showed that missing data can cause problems for recovering ``correct'' topologies, especially for small matrices of 100 characters (Wiens 2003).
It is important to note, however, that this effect can be reduced by increasing the number of characters (Wiens 2003).} lines 632-644. @@@

% -----------------
% Comment 11
% -----------------
\item{\textcolor{blue}{\textbf{Conclusion:} The authors should clarify how their results compare to those of previous studies more directly to emphasize the nuances that make this study unique. }}

We rewrote the conclusion to deal with this comment (and the specific comments 13, 52 - see comment 52 for details on the choice of citing Arcila et al 2015. - and 56), as follows:

% TG: I rephrased as follows.
\textit{Previous studies have explored the effect of missing morphological data in Total Evidence matrices (Wiens et al., 2005; Manos et al.,2007, Pattinson et al., 2014).
The conclusions of theses studies, however, were limited by their empirical approach making their results applicable only to similar scenarios. %TG: or scenari
Additionally, these studies focused either only on living taxa (Wiens et al., 2005) or on the patterns of missing data from the fossil record only (Manos et al.,2007, Pattinson et al., 2014).
The conclusions of these studies were therefore restricted to these specific aspect of missing data and, besides their interesting conclusions, % TG: seems unfair to me to just bash them, they are actually nice papers
they didn't provide any insight of the combined effect of different missing data parameters.
% as well as several aspects of their experimental design (e.g. not including fossil taxa, Wiens et al., 2005; or focusing on the palaeontological aspect, Manos et al.,2007, Pattinson et al., 2014).
% NC: Again needs a bit more explanation. WHY is empirical approach a problem? Because it only gives answers to one specific scenario or clade?
Here we used a more in depth approach where missing data were generated from simulated data and according to three clearly defined missing-data parameters ($M_{L}$, $M_{F}$ or $N_{C}$).
This allowed us to confirm previous results that missing data can be especially problematic in small matrices (Wiens 2003), but also revealed the crucial importance of coding morphological data for living species in Total Evidence phylogenies.
Missing data in Total Evidence matrices is not a problem for recovering the ``best'' tree topology as long as enough living and fossil taxa in the matrix have data for overlapping morphological characters.
When missing data increases in any of our missing data parameters ($M_{L}$, $M_{F}$ or $N_{C}$), it reduces support for the placement of fossil taxa and increases the displacement of wildcard taxa.
Therefore we advise increased focus on coding morphological characters for a large number of the living taxa present in the matrix (i.e. 50\%) if the goal is to accurately combine both living and fossil species in phylogenies.
Doing so will increase overlap of morphological characters among living and fossil taxa, allowing the fossil taxa to be positioned relative to the living taxa based on their shared derived characters rather than simply on available data.} lines @@@

\textit{Additionally, the topology of the Bayesian consensus trees, regardless of the amount of missing data, were always closer to the ``best'' tree topology than the Maximum Likelihood trees.
This has also been observed in empirical data (e.g. Arcila et al., 2015) where Maximum Likelihood trees inferred from a Total Evidence matrix were less supported than the Bayesian consensus tree.
This might have an important impact on estimating topologies in the Total Evidence framework, because previous studies had to rely either on molecular scaffolds (e.g. Slater, 2013), taxonomic constraints (e.g. Slater, 2013; Beck and Lee, 2014) or even on fixing the topology (e.g. Ronquist et al., 2012a).
Therefore, we suggest extracting such topological backbones from the Bayesian consensus tree if needed.} lines @@@
% TG: maybe add in a non tip-dating analysis way (i.e. Nick Matzke comment)? 
% NC: Sure if it fits. If you want to "cite" Nick you can use (N.Matzke pers.comm.)
% TG: ok, actually leave it, I've tried something but the cite wasn't that convincing and we're already being carefull enough here.

% \textit{To conclude, the results of our analyses are encouraging and show that it is possible to accurately combine both neontological and palaeontological morphological data in the same phylogeny as long as there is sufficient overlap.
% Hopefully, using these approaches will greatly improve our understanding of macroevolutionary patterns and processes.} lines 602-634. @@@
% NC: I don't like this last bit. It's really weak. I'd cut it to be honest. The first sentence is not very clear (suddenly you revert back to neo and paleo) and the second sentence has no real explanation or link to the rest of the paper. TG: let's just remove it then... You're right it's totally useless and after reading the whole paper, the reader won't mind saving a few more second of their life.

@@

We changed the axes of the graphs showing Normalised Triplets metric to show values below 0 throughout the manuscript and the appendices.

% -----------------
% Comment 14
% -----------------
\item{\textcolor{blue}{\textbf{Table caption:} Table caption should include a line about 'pooling across missing data schemes' to clarify what is presented.}}

We clarified the caption of the Table 1 as follows:

\textit{Each line summarizes the probabilities of overlap between the distributions of the ``best'' tree versus trees from each inference method (Maximum Likelihood; Bayesian consensus; Maximum Likelihood Bootstraps and Bayesian posterior trees) pooled across all combinations of missing data parameter values, using the Normalised Robins-Foulds (RF) and Triplets (Tr) metrics. 
Values highlighted in bold are the extreme values of high or low probability of overlap between two methods. If two methods have a high probability of overlap, they have a similar ability to recover the ``correct'' tree topology. Values $>$ 0.95 denote significantly similar distributions and values $<$ 0.05 denote significantly different distributions."}
% NC: Do you want to say values > 0.95 are significantly similar, values < 0.05 are significantly different?  TG: yes.

@@

(1) - We thank the reviewer for this comment and now emphasize both the novel aspects of our study and the questions we are actually testing (see reviewer 1 comment 4 above).

(2) - the second point underlined by the reviewer in this specific comment strikes use as really interesting.
We ran this additional analysis by comparing the estimated $\alpha$ parameter for the morphological rates distribution to the ``true'' $\alpha$ parameter.
We found results that the $\alpha$ parameter is usually slightly underestimated for the various ``missing-data'' trees unless the number of living species with no morphological data is high ($M_L$ = 75\%), then, the $\alpha$ are overestimated.
These results are somewhat in agreement with our topological differences results: the number of living species seems to be a driving missing data parameter for inferring Total Evidence trees.
Nonetheless, we considered this study to already be long and complex.
We therefore think that adding an extra question to this study (i.e. what is the effect of missing data on the morphological substitution rate parameter estimates?) might lose the reader and make an already long paper even longer.
Therefore, we added the results of these analyses to \textbf{Appendix-A Morphological rates estimations results}.
We link to this analysis in the methods as follows:

\textit{The detailed MrBayes parameters are available in Appendix A. We also included analysis showing details on the $\alpha$ parameter estimation though it is beyond the scope of this paper and are not discussed further here.} lines 336-338 @@@
See also Appendix A "Morphological rates estimations results".

% NC: Needs a bit more explanation. Something like, we also include in the appendix an analysis showing ... though it is beyond the scope of this paper to discuss these results further here. TG: is that ok?

@@

2) We followed the reviewer's suggestion and added six extra tables to Appendix C (Table C8 to C13) containing the marginal probability of each missing data parameter individually and for each tree inference method.

% -----------------
% Specific Comment 59
% -----------------
\item{\textcolor{blue}{\textbf{Appendix C:} Tables 1-3 repeats 'of the'. Should label the first two rows as 'Best maximum likelihood tree' and 'Best Bayesian inference tree'}}

We removed the repetition in the caption and changed the first two rows as suggested.
%If we understood the second point correctly, the reviewer suggests we add ``\textit{best}'' to the two first rows of the tables, and by ``\textit{best}'', the reviewer means ``optimal'' (as opposed to the ``best'' tree in our protocol that refers to any tree inferred by any method from the ``complete matrix''.
% We don't think it is necessary to follow this suggestion as the Maximum Likelihood tree is by definition the tree with the highest likelihood, and the Bayesian consensus is the consensus tree of the Bayesian inference trees.
%So adding the word ``\textit{best}'' here is not entirely accurate.

% NC: No we discussed this last night - I think they *do* mean "best" as in our protocol. So just make the changes to the first two rows. TG: Erf, deeply sorry, I probably forgot to take the not here while chatting. Ok after reflection I either don't remember at all what I said or either don't agree. The tables show the "Summary of the comparisons between the "best" tree and the "missing-data" trees for each different tree inference method using either the Normalised Robinson-Foulds metric (RF) or the Normalised Triplets metric (Tr)." 
%
% Tree inference method       Metric   Min.  1st Qu.  etc...
% Maximum Likelihood tree       RF      1.1     2.2     3.3
%
% And this reads: the comparison between the "best" <insert what's in the Tree inference method column> tree compared to the "missing-data" trees in the same method for the <insert metric> is minimum <insert value> etc...

@@

\section{Reviewer 2:}
Note that we split the second comment from reviewer 2 in three different comments (comments 2,3 and 4 below) to improve the readability of our response to the reviewers.

% -----------------
% Comment 1
% -----------------
\begin{enumerate}
\item{\textcolor{blue}{\textbf{Compare the trees to the "true" tree:} I don't see how the "best" tree approach is helpful in this case. It adds an additional layer of complexity that can't be accounted for via simulation. The point of simulations is to compare empirical realities to the generating process, to see how much results diverge. I think the most helpful and straightforward thing to do is eliminate the "best" tree part of the MS, and compare the inferred trees directly to the real topology to quantify variance.}}

A similar point/suggestion was also raised by reviewer 1 and we ran the suggested extra analysis. See reviewer 1 general comment number 6 and specific comment number 25 for a detailed response.

We understand the concerns of the reviewer and added the comparisons between the ``true'' and the ``missing-data'' trees to the Appendix C.
Nonetheless, we would prefer to stick with our "best" tree approach in the main manuscript.
In practice, the ``true'' tree can be perceived as a ``random'' seed to generate the ``complete'' matrix, as it simply a birth-death tree with $\lambda$ $>$ $\mu$ and a set number of living and fossil taxa (see details in the manuscript section "2.1 Generating the matrix").
All these parameters were only used to create a realistic ``complete'' matrix in order to make our results relevant to biologists.

% NC: Can you rephrase a little here so it sounds less like you saying "no, and you're stupid". Doesn't the response to reviewer 1 cover this already? I've had a go but you may need to check what I've written. TG: I've added the bit at the start saying "no but look, we did it in the appendix". I didn't really get the "and you're stupid" (I'm not the best person to judge these things ;)). Maybe add something along the line "And we're sorry for the confusion, we now the manuscript is long and complicated but we made a effort of describing each step in great details, even the random seed generation..."

### 2015-08-11 Manuscript/MPESubmission/Revision/response.tex

% -----------------
% Comment 2
% -----------------
\item{\textcolor{blue}{\textbf{Better explanation of the missing data:} I found the explanation of the missing data strategy very difficult to follow. First, they oddly use the verb "to infer" for matrices; are they inferring matrices? Second, it's not clear anywhere that the authors actually removed the molecular data from the fossil taxa, though I assume they did. I hope the authors can re-write in simpler language ("explain like I'm 5") for easier comprehension.}}

The lines 180 and 181 states "All the molecular information for fossil taxa was replaced by missing data ("?")" in the former manuscript. We changed the verb "infer" (see specific comment number 16 from reviewer 1).

We assume the reviewer is referring to section "2.2 Removing data". We therefore modified the whole section to improve its clarity and simplicity as follows:

\textit{To explore the effect of missing morphological data on topological recovery, we removed various amounts of the ``complete'' matrix to obtain matrices with missing morphological data.
Hereafter, we call these matrices with missing morphological data the ``missing-data'' matrices.
Note that the amount of molecular data remained constant throughout our simulations: 1000 molecular characters for living taxa and no molecular data for fossil taxa (see above).
We removed morphological data using three data incompleteness parameters:
\begin{enumerate}
\item{The proportion of missing living taxa ($M_L$).} This first missing-data parameter corresponds to the proportion of living taxa with no morphological data. It represents the number of living taxa that are present in the matrix but have only molecular data available. This reflects the fact that, because of the increasing ease of collecting molecular data, morphological data for living species is rarely collected (Guillerme and Cooper, 2015). Therefore, many living species will have only molecular data available. In practice, we removed all the morphological data from randomly chosen living taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of living taxa with no morphological data.
\item{The proportion of missing data in the fossil record ($M_F$).} This missing data parameter represents the completeness of the fossil record. Due to preservation biases, missing data for fossil taxa are common (Sansom and Wills, 2013). In practice, we randomly removed a proportion of data among the fossil taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of overall missing data for the fossil taxa.
%Added this part following comment 4.
Note that 50\% missing data for fossil taxa does not mean that each fossil is missing 50\% of its morphological data.
Instead this 50\% refers to missing fossil data across the whole matrix.
Some fossils may retain 100\% of their data and others may lose all their data at this parameter value (down to a minimum of 5\% available data; see below)
%up to a maximum of 95\% missing data, see below).
% NC: normally you'd say up to a maximum rather than up to a minimum. Is what I've written correct? TG: maybe change it "down to a minimum"?
\item{The number of morphological characters for both living and fossil taxa ($N_C$).} This parameter is not a missing data parameter \textit{per se} but rather an indication of the size of the matrix. Any morphological matrix of any size has indeterminate missing data, given that the total number of characters is undefined, but presumably large. Therefore, this parameter corresponds to the overall number of characters available for both living and fossil taxa. In practice, we randomly removed whole characters from the morphological matrix reducing it to: 100, 90,75, 50 or 25 characters.
Note that these levels are equivalent to the two other parameters (i.e. 0\%, 10\%, 25\%, 50\% or 75\% of ``missing'' morphological characters).
\end{enumerate}}
